An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-09 paper.tex
Section 2 fixes.
Edits to Ilya's Section V, re-formatting.

We have successfully employed the RJMCMC technique described above when evaluating several alternative models for the distribution of the masses of black holes in X-ray binaries \cite{Farr:2010}.  We performed a Bayesian analysis of the mass distribution of stellar-mass black holes using the observed masses of 15 low-mass X-ray binary systems undergoing Roche lobe overflow and five high-mass, wind-fed X-ray binary systems.  We considered ten different mass distribution models: Gaussian, double Gaussian, power law, exponential decay, log normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was described by between two and six parameters.  Model selection using RJMCMC with kD trees allowed us to determine that the mass distribution of the low-mass systems is  best fit by a power-law, while the distribution of the combined sample is best fit by the exponential model.  We were thus able to conclude that the low-mass subsample is not consistent with being drawn from the distribution of the combined population \cite{Farr:2010}.

Gravitational-wave astronomy provides the setting for another ongoing study.  LIGO and Virgo gravitational-wave detectors have recently completed taking data, which is currently being analyzed.  Interesting triggers from the detection pipeline that  searches for compact binary coalescences are followed up with Bayesian parameter estimation tools.  In general, several possible models for the data are considered: it can contain pure noise, or noise superimposed on a gravitational-wave signal that may or may not include significant spins in the binary components.  The models can include a large number of parameters (up to 15 for the model with two spinning components), and RJMCMC with interpolation via kD trees is again successful at computing the model evidences. \ilya{Add references throughout this paragraph.}

@@

The possibility of MCMC chains becoming even temporarily stuck on local maxima also affects convergence.   In the absence of a sufficient number of jump proposals between the local maxima, the MCMC may not accurately represent their relative sizes.  For example, this can happen with antipodal sky location degeneracies for networks of three ground-based gravitational-wave interferometers or for low-frequency gravitational wave signals in the proposed space-borne instrument LISA \cite{LISA}.  Although PDFs produced by MCMC chains will generally show both locations on the sky, the chain length may not be sufficiently long in practice to include enough jumps between the two high-likelihood antipodal locations.  We propose that the convergence of such PDFs can be tested by interpolating them with kD trees as described above and using them as jump proposal distributions in a follow-up MCMC.  This follow-up MCMC would make large global jump proposals rather than small local ones, meaning that it could rapidly accumulate sufficient points to determine whether the original run spent correct fractional amounts of time in the various modes.

\ilya{Mention nested sampling?}

>>>>

We have successfully employed the RJMCMC technique described above
when evaluating several alternative models for the distribution of the
masses of black holes in X-ray binaries \cite{Farr:2010}.  We
performed a Bayesian analysis of the mass distribution of stellar-mass
black holes using the observed masses of 15 low-mass X-ray binary
systems undergoing Roche lobe overflow and five high-mass, wind-fed
X-ray binary systems.  We considered ten different mass distribution
models: Gaussian, double Gaussian, power law, exponential decay, log
normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was
described by between two and six parameters.  Model selection using
RJMCMC with kD trees allowed us to determine that the mass
distribution of the low-mass systems is best fit by a power-law, while
the distribution of the combined sample is best fit by the exponential
model.  Based on the model selection, we were able to determine which
models would provide the best information about astrophysically
relevant parameters, such as the minimum black hole mass.  We were
also able to conclude that the low-mass subsample is not consistent
with being drawn from the distribution of the combined population
\cite{Farr:2010}.

Gravitational-wave astronomy provides the setting for another ongoing
study.  Interesting triggers from the LIGO and Virgo
gravitational-wave detector pipeline that searches for compact binary
coalescences are followed up with Bayesian parameter estimation tools.
In general, several possible models for the data are considered,
including a pure noise model, noise superimposed on a non-spinning
gravitational-wave signal, or noise superimposed on a gravitational
wave signal that includes significant spin or spins in the binary
components.  The models can include a large number of parameters (up
to 15 for the model with two spinning components), and RJMCMC with
interpolation via kD trees is again successful at computing the model
evidences. \ilya{Add references throughout this paragraph.}

@@

The possibility of MCMC chains becoming even temporarily stuck on
local maxima also affects convergence.  In the absence of a sufficient
number of jump proposals between the local maxima, the MCMC may not
accurately represent their relative sizes.  For example, this can
happen with antipodal sky location degeneracies for networks of three
ground-based gravitational-wave interferometers or for low-frequency
gravitational wave signals in the proposed space-borne instrument LISA
\cite{LISA}.  Although PDFs produced by MCMC chains will generally
show both locations on the sky, the chain length may not be
sufficiently long in practice to include enough jumps between the two
high-likelihood antipodal locations.  In general, the relative weight
of two separated peaks is determined with a fractional error that
scales as $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of transitions between
the peaks.  By increasing the number of proposed transitions, the
convergence of such PDFs can be enhanced using the kD tree
interpolated proposal.  We can test the convergence of an MCMC with
local jumps by running a followup MCMC with a jump proposal
distribution from an interpolation of the first MCMC's PDF.  This
follow-up MCMC would make large global jump proposals rather than
small local ones, meaning that it could rapidly accumulate sufficient
points to determine whether the original run spent correct fractional
amounts of time in the various modes.
